---
title: "Syllabus"
---

```{r include = FALSE}
library(readxl)
library(dplyr)
library(lubridate)
library(stringr)
```

```{r include = FALSE}
# import excel table with topics
rawdat <- read_excel(file.path("course-admin", "schedule.xlsx"))

# clean up table and extract date info
cdat <- rawdat %>%
  mutate(wday = wday(date, label = TRUE),
         month = month(date, label = TRUE),
         day = day(date),
         date = paste(wday, month, day),
         date = ifelse(is.na(wday), NA, date),
         link_it = as.logical(link_it),
         notes = sprintf("%s: %s", cm, topic))

# identify links to course modules
cmdat <- data_frame(html = list.files(pattern = '^(cm|setup|git).*html$'))
cmdat <- cmdat %>% mutate(cm = str_extract(html, "(cm|setup|git)[0-9]+"))
cdat <- cdat %>% left_join(cmdat)

# add links to notes column
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it],
              "((cm[0-9]{3})|((setup|git)[0-9]{2}))", '<a href="LINK">\\1</a>')
cdat$notes[cdat$link_it] <-
  str_replace(cdat$notes[cdat$link_it], "LINK", cdat$html[cdat$link_it])
```

```{r results = 'asis', echo = FALSE}
cdat %>%
  select(date, notes) %>%
  knitr::kable(col.names = c("Date", "Topic"))
```

## Homework

* [Homework 01/02: Wrangling and exploring data](hw02-wrangle-explore-data.html)
* [Homework 03: Probability](hw03-probability.html)
* [Homework 04: Discrete random variables](hw04-rv-discrete.html)
* [Homework 05: Continuous random variables and optimization](hw05-rv-continuous-optimize.html)
* [Homework 06: Statistical inference, parametric models, and hypothesis testing](hw06-stat-inference.html)
* [Homework 07: Bayesian inference and OLS estimator](hw07-bayesian-ols.html)
