---
title: "Graphical data analysis"
author: |
  | MACS 33001
  | University of Chicago
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(ggthemes)
library(knitr)
library(broom)
library(stringr)
library(patchwork)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal(base_size = 18))
```

## Graphical data analysis

1. Generate questions about your data
1. Search for answers by visualising, transforming, and modeling your data
1. Use what you learn to refine your questions and or generate new questions
1. Rinse and repeat until you publish a paper

## Graphical data analysis

1. What type of variation occurs within my variables?
1. What type of covariation occurs between my variables?
1. Are there outliers in the data?
1. Do I have missingness? Are there patterns to it?
1. How much variation/error exists in my statistical estimates? Is there a pattern to it?

## Differences between GDA and modeling

Tips dataset

Variable | Explanation
---------|-------------
`obs` | Observation number
`totbill` | Total bill (cost of the meal), including tax, in US dollars
`tip` | Tip (gratuity) in US dollars
`sex` | Sex of person paying for the meal (0=male, 1=female)
`smoker` | Smoker in party? (0=No, 1=Yes)
`day` | 3=Thur, 4=Fri, 5=Sat, 6=Sun
`time` | 0=Day, 1=Night
`size` | Size of the party

```{r get-tips, include = FALSE}
tips <- read_csv("../data/tips.csv")
str(tips)
```

## Tips regression

```{r tips-lm}
tips <- tips %>%
  mutate(tiprate = tip / totbill)

tips_lm <- lm(tiprate ~ sex + smoker + day + time + size, data = tips)
tidy(tips_lm)
```

```{r tips-rsq}
glance(tips_lm)
```

## Exploring `tips`

```{r tips-hist}
ggplot(tips, aes(tip)) +
  geom_histogram(binwidth = 1)
```

## Exploring `tips`

```{r tips-round}
ggplot(tips, aes(tip)) +
  geom_histogram(binwidth = .1)
```

## Exploring `tips`

```{r tips-scatter}
ggplot(tips, aes(totbill, tip)) +
  geom_point() +
  geom_abline(slope = .18, linetype = 2) +
  expand_limits(x = c(0, NA),
                y = c(0, NA)) +
  geom_text(data = tips %>%
              summarize(rsq = format(cor(totbill, tip), digits = 3)),
            aes(x = 2, y = 9, label = rsq))
```

## Exploring `tips`

```{r tips-scatter-many}
ggplot(tips, aes(totbill, tip)) +
  geom_point() +
  geom_abline(slope = .18, linetype = 2) +
  expand_limits(x = c(0, NA),
                y = c(0, NA)) +
  facet_grid(smoker ~ sex, labeller = "label_both") +
  geom_text(data = tips %>%
              group_by(smoker, sex) %>%
              summarize(rsq = format(cor(totbill, tip), digits = 3)),
            aes(x = 2, y = 9, label = rsq))
```

## GDA vs. CDA

* Graphical data analysis
* Confirmatory data analysis

## Histograms {.scrollable}

```{r infant-data, include = FALSE}
infant <- read_csv("../data//infant.csv") %>%
  # remove non-countries
  filter(is.na(`Value Footnotes`) | `Value Footnotes` != 1) %>%
  select(`Country or Area`, Year, Value) %>%
  rename(country = `Country or Area`,
         year = Year,
         mortal = Value)
```

```{r infant-hist}
ggplot(infant, aes(mortal)) +
  geom_histogram(bins = 10, boundary = 0) +
  labs(title = "Histogram of infant mortality rate for 195 nations",
       subtitle = "10 bins, origin = 0",
       x = "Infant mortality rate (per 1,000)",
       y = "Frequency")

ggplot(infant, aes(mortal)) +
  geom_histogram(bins = 10, boundary = -5) +
  labs(title = "Histogram of infant mortality rate for 195 nations",
       subtitle = "10 bins, origin = -5",
       x = "Infant mortality rate (per 1,000)",
       y = "Frequency")
```

## Nonparametric density estimation

$$x_0 + 2(j - 1)h \leq X_i < x_0 + 2jh$$

$$\hat{p}(x) = \frac{\#_{i = 1}^n [x_0 + 2(j - 1)h \leq X_i < x_0 + 2jh]}{2nh}$$

$$\hat{p}(x) = \frac{\#_{i = 1}^n [x-h \leq X_i < x+h]}{2nh}$$

$$\hat{p}(x) = \frac{1}{nh} \sum_{i = 1}^n W \left( \frac{x - X_i}{h} \right)$$

$$W(z) = \begin{cases} 
  \frac{1}{2} & \text{for } |z| < 1 \\
  0 & \text{otherwise} \\
  \end{cases}$$

$$z = \frac{x - X_i}{h}$$

## Naive density estimation

```{r}
ggplot(infant, aes(mortal)) +
  geom_density(kernel = "rectangular") +
  labs(title = "Naive density estimator of infant mortality rate for 195 nations",
       x = "Infant mortality rate (per 1,000)",
       y = "Density")
```

## Kernels

$$\hat{x}(x) = \frac{1}{nh} \sum_{i = 1}^k K \left( \frac{x - X_i}{h} \right)$$

## Gaussian kernel {.scrollable}

$$K(z) = \frac{1}{\sqrt{2 \pi}}e^{-\frac{1}{2} z^2}$$

```{r gaussian}
x <- rnorm(1000)

{
  qplot(x, geom = "blank") +
    stat_function(fun = dnorm) +
    labs(title = "Gaussian (normal) kernel",
         x = NULL,
         y = NULL)
} +
{
  ggplot(infant, aes(mortal)) +
    geom_density(kernel = "gaussian") +
    labs(title = "Infant mortality rate for 195 nations",
         x = "Infant mortality rate (per 1,000)",
         y = "Density")
}
```

## Comparison of kernels

```{r kernels}
triangular <- function(x) {
  (1 - abs(x)) * ifelse(abs(x) <= 1, 1, 0)
}

biweight <- function(x) {
  (15 / 16) * (1 - x^2)^2 * ifelse(abs(x) <= 1, 1, 0)
}

epanechnikov <- function(x) {
  (15 / 16) * (1 - x^2)^2 * ifelse(abs(x) <= 1, 1, 0)
}

library(gridExtra)
kernels <- qplot(x, geom = "blank") +
    stat_function(aes(color = "Gaussian"), fun = dnorm) +
    stat_function(aes(color = "Epanechnikov"), fun = epanechnikov) +
    stat_function(aes(color = "Rectangular"), fun = dunif,
                  args = list(min = -1), geom = "step") +
    stat_function(aes(color = "Triangular"), fun = triangular) +
    stat_function(aes(color = "Biweight"), fun = biweight) +
    scale_color_brewer(type = "qual") +
    labs(x = NULL,
         y = NULL,
         color = NULL) +
    theme(legend.position = "bottom",
          legend.background = element_rect(fill = "white"))

kdes <- ggplot(infant, aes(mortal)) +
    geom_density(aes(color = "Gaussian"), kernel = "gaussian") +
    geom_density(aes(color = "Epanechnikov"), kernel = "epanechnikov") +
    geom_density(aes(color = "Rectangular"), kernel = "rectangular") +
    geom_density(aes(color = "Triangular"), kernel = "triangular") +
    geom_density(aes(color = "Biweight"), kernel = "biweight") +
    scale_color_brewer(type = "qual") +
    labs(title = "Density estimators of infant mortality rate for 195 nations",
         x = "Infant mortality rate (per 1,000)",
         y = "Density",
         color = "Kernel") +
    theme(legend.background = element_rect(fill = "white"))

g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
  }

mylegend <- g_legend(kernels)

p3 <- grid.arrange(arrangeGrob(kernels + theme(legend.position="none"),
                         kdes + theme(legend.position="none"),
                         nrow=1),
             mylegend, nrow = 2, heights = c(10, 1))
```

## Selecting the bandwidth $h$

```{r gaussian-h}
ggplot(infant, aes(mortal)) +
  geom_density(kernel = "gaussian", adjust = 5) +
  geom_density(kernel = "gaussian", adjust = 1, linetype = 2) +
  geom_density(kernel = "gaussian", adjust = 1/5, linetype = 3) +
  labs(title = "Gaussian density estimators of infant mortality rate for 195 nations",
       subtitle = "Three different bandwidth parameters",
       x = "Infant mortality rate (per 1,000)",
       y = "Density")
```

## Boxplot

```{r boxplot}
ggplot(infant, aes("", mortal)) +
  geom_boxplot() +
  labs(title = "Boxplot of infant mortality rate for 195 nations",
       x = NULL,
       y = "Infant mortality rate (per 1,000)")
```

## Violin plot

```{r violin}
ggplot(infant, aes("", mortal)) +
  geom_violin() +
  labs(title = "Violin plot of infant mortality rate for 195 nations",
       x = NULL,
       y = "Infant mortality rate (per 1,000)")
```

## Things to look for in continuous variables

* Assymetry
* Outliers
* Multimodality
* Gaps
* Heaping
* Rounding
* Impossibilities
* Errors

## Galton's heights

```{r get-galton}
data(galton, package = "UsingR")

galton_tidy <-  as_tibble(galton) %>%
  gather(person, height)

ggplot(galton_tidy, aes(height)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ person, scales = "free")
```

## Investigate for gaps or heaping

```{r galton-heaping}
ggplot(galton_tidy, aes(height)) +
  geom_histogram(binwidth = .1) +
  facet_wrap(~ person, scales = "free")
```

## Comparing the distributions

```{r galton-both}
ggplot(galton_tidy, aes(height, fill = person)) +
  geom_histogram()
```

## Comparing the distributions

```{r galton-sep}
galton_median <- galton_tidy %>%
  group_by(person) %>%
  summarize(median = median(height))

ggplot(galton_tidy) +
  geom_histogram(aes(height), binwidth = 1) +
  geom_vline(data = galton_median, aes(xintercept = median), color = "red") +
  facet_wrap(~ person, ncol = 1)
```

## Comparing the distributions {.scrollable}

```{r galton-density}
ggplot(galton_tidy, aes(height, color = person)) +
  geom_density()
```

```{r galton-box}
ggplot(galton_tidy, aes(person, height)) +
  geom_boxplot()
```

## Outlier detection

```{r movies-hist}
data(movies, package = "ggplot2movies")

ggplot(movies, aes(length)) +
  geom_histogram()
```

## Outlier detection

```{r movies-box}
ggplot(movies, aes("", length)) +
  geom_boxplot() +
  labs(x = NULL)
```

## Outlier detection

```{r movies-filter}
filter(movies, length > 1000)
```

## Filter outliers {.scrollable}

```{r movies-all}
movies_lite <- filter(movies, length <= 180)

ggplot(movies_lite, aes("", length)) +
  geom_boxplot() +
  coord_flip() +
  labs(x = NULL,
       y = "Movie length (in minutes)")

ggplot(movies_lite, aes(length)) +
  geom_density() +
  labs(x = "Movie length (in minutes)")

ggplot(movies_lite, aes(length)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Movie length (in minutes)")
```

## Compare distributions of subgroups

```{r galton-density2}
ggplot(galton_tidy, aes(height, color = person)) +
  geom_density()
```

## Compare distributions of subgroups

```{r movies-genre}
genre <- movies_lite %>%
  gather(genre, value, Action:Short) %>%
  filter(value == 1) %>%
  select(-value)
```

```{r genre-density}
ggplot(genre, aes(length, color = genre)) +
  geom_density() +
  scale_color_brewer(type = "qual")
```

## Multiple windows plot

```{r genre-facet}
ggplot(genre, aes(length)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Movie length (in minutes)") +
  facet_wrap(~ genre)
```

## Multiple windows plot

```{r genre-facet-freey}
ggplot(genre, aes(length)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Movie length (in minutes)") +
  facet_wrap(~ genre, scales = "free_y")
```

## Boxplot

```{r genre-box}
ggplot(genre, aes(genre, length)) +
  geom_boxplot() +
  labs(x = "Genre",
       y = "Movie length (in minutes)")
```

## Categorical variables

* Discrete variables with a fixed set of possible values

## Bar chart

```{r gss_cat, include = FALSE}
library(forcats)
gss_cat
```

```{r race-bar}
ggplot(gss_cat, aes(race)) +
  geom_bar()
```

## Omitted categories

```{r race-bar-omit}
ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

## Order matters

```{r relig-bar}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(relig, tvhours)) +
  geom_col() +
  coord_flip()
```

## Order matters

```{r relig-alpha}
ggplot(relig_summary, aes(factor(relig, levels = rev(sort(levels(relig)))), tvhours)) +
  geom_col() +
  coord_flip()
```

## Order matters

```{r relig-order}
ggplot(relig_summary, aes(fct_reorder(relig, tvhours), tvhours)) +
  geom_col() +
  coord_flip()
```

## Order matters

```{r income-plot}
rincome_summary <- gss_cat %>%
  group_by(rincome) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(rincome_summary, aes(age, fct_reorder(rincome, age))) +
  geom_point()
```

## Order matters

```{r income-bottom}
ggplot(rincome_summary, aes(age, fct_relevel(rincome, "Not applicable"))) +
  geom_point()
```

## Variations on bar charts

```{r intent}
data(gun_deaths, package = "rcfss")

gun_deaths %>%
  count(intent) %>%
  na.omit() %>%
  ggplot(aes(fct_reorder(intent, n, .desc = TRUE), n)) +
  geom_col()
```

## Stacked bar chart

```{r intent-stack}
gun_deaths %>%
  count(intent, race) %>%
  na.omit() %>%
  mutate(intent = factor(intent, levels = c("Suicide", "Homicide",
                                            "Accidental", "Undetermined"))) %>%
  ggplot(aes(intent, n, fill = race)) +
  geom_col()
```

## Dodged bar chart

```{r intent-dodge}
gun_deaths %>%
  count(intent, race) %>%
  na.omit() %>%
  mutate(intent = factor(intent, levels = c("Suicide", "Homicide",
                                            "Accidental", "Undetermined"))) %>%
  ggplot(aes(intent, n, fill = race)) +
  geom_col(position = "dodge")
```

## Proportional bar chart

```{r intent-fill}
gun_deaths %>%
  count(intent, race) %>%
  na.omit() %>%
  mutate(intent = factor(intent, levels = c("Suicide", "Homicide",
                                            "Accidental", "Undetermined"))) %>%
  ggplot(aes(intent, n, fill = race)) +
  geom_col(position = "fill")
```

## Scatterplots

* Causal relationships (linear and nonlinear)
* Associations (correlations)
* Outliers or groups of outliers
* Clusters
* Gaps
* Barriers
* Conditional relationships

## `movies` example

```{r movies-votes-rating}
ggplot(movies, aes(votes, rating)) +
  geom_point()
```

## Smoothing lines

```{r movies-votes-rating-smooth}
ggplot(movies, aes(votes, rating)) +
  geom_point() +
  geom_smooth()
```

## Adding jitter to the graph

```{r heights-scatter}
ggplot(galton, aes(parent, child)) +
  geom_point()
```

## Adding jitter to the graph

```{r heights-jitter}
ggplot(galton, aes(parent, child)) +
  geom_jitter()
```

## Adding jitter to the graph

```{r galton-smooth}
ggplot(galton, aes(parent, child)) +
  geom_jitter(alpha = .2) +
  geom_abline(slope = 1, linetype = 2) +
  geom_smooth(method = "lm")
```

## Comparing groups within scatterplots

```{r auto-drv}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point()
```

## Comparing groups within scatterplots

```{r auto-class}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Comparing groups within scatterplots

```{r olympics}
data(oly12, package = "VGAMdata")

ggplot(oly12, aes(Height, Weight, color = Sport)) +
  geom_point()
```

## Comparing groups within scatterplots

```{r olympics-facet}
oly12 %>%
  mutate(Sport = abbreviate(Sport, 12)) %>%
  ggplot(aes(Height, Weight)) +
  geom_point(size = 1) +
  facet_wrap(~ Sport)
```

## Comparing groups within scatterplots

```{r olympics-facet-lite}
oly12 %>%
  filter(Sport %in% c("Judo", "Weightlifting", "Wrestling")) %>%
  ggplot(aes(Height, Weight)) +
  geom_point(size = 1) +
  facet_wrap(~ Sport)
```

## Scatterplot matrix

```{r credit-import}
credit <- read_csv("../data/Credit.csv") %>%
  # remove first ID column
  select(-X1)
names(credit) <- stringr::str_to_lower(names(credit))   # convert column names to lowercase
```

```{r credit-scatter-matrix}
# only graph numeric variables
pairs(select_if(credit, is.numeric))
```

## Scatterplot matrix

```{r credit-scatter-ggpairs}
library(GGally)

ggpairs(select_if(credit, is.numeric))
```

## Scatterplot matrix

```{r diamonds-scatter-matrix, warning = FALSE, message = FALSE}
ggpairs(select(rcfss::scorecard, type:debt))
```

## Heatmap of correlation coefficients {.scrollable}

```{r ggcorr}
ggcorr(mpg)
ggcorr(credit)
ggcorr(diamonds)
```

## Heatmap of correlation coefficients {.scrollable}

```{r ggcorr-label}
ggcorr(credit, label = TRUE)
ggcorr(credit, label = TRUE, label_round = 2, label_alpha = TRUE)
```

