---
title: "Functions and iterative operations"
author: |
  | MACS 33001
  | University of Chicago
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(patchwork)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal(base_size = 18))
```

## Functions

* Easy to reuse
* Self-documenting
* Easy-ier to debug
* [If you have copied and pasted a block of code more than twice, convert it to a function.](http://r4ds.had.co.nz/functions.html#when-should-you-write-a-function)

## Function components

* Name
* Arguments
* Body

## Rescale function

```{r, collapse = TRUE}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale01(c(0, 5, 10))
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

* Name
* Arguments
* Body

## What is that?

```{r}
pythagorean <- function(a, b){
  hypotenuse <- sqrt(a^2 + b^2)
  return(hypotenuse)
}
```

* Name
* Arguments
* Body

## How to use a function {.scrollable}

```{r, error = TRUE}
# print the output of the function
pythagorean(a = 3, b = 4)

# save the output as a new object
(tri_c <- pythagorean(a = 3, b = 4))

# what happens to the hypotenuse from inside the function?
pythagorean(a = 3, b = 4)
hypotenuse
```

## Iteration

```{r df-rnorm}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

```{r median-per-column, collapse = TRUE}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

## Iteration with `for` loop

```{r for-loop}
output <- vector(mode = "double", length = ncol(df))
for (i in seq_along(df)) {
  output[[i]] <- median(df[[i]])
}
output
```

## Output {.scrollable}

```r
output <- vector(mode = "double", length = ncol(df))
```

```{r vector}
vector(mode = "double", length = ncol(df))
vector(mode = "logical", length = ncol(df))
vector(mode = "character", length = ncol(df))
vector(mode = "list", length = ncol(df))
```

## Sequence

```r
i in seq_along(df)
```

```{r seq-along}
seq_along(df)
```

## Body

```r
output[[i]] <- median(df[[i]])
```

## Preallocation {.scrollable}

```{r preallocate, message = FALSE}
x <- rnorm(1000, mean = 0, sd = 1)
str(x)

# load microbenchmark library to time code
library(microbenchmark)

microbenchmark(
  # don't preallocate
  `No preallocation` = {
    output <- vector("numeric", 0)
    
    for (i in seq_along(x)) {
      output <- c(output, x[[i]] + 1)
    }
  },
  # preallocate
  `Preallocation` = {
    output <- vector("numeric", length(x))
    
    for (i in seq_along(x)) {
      output[[i]] <- x[[i]] + 1
    }
  }) %>%
  autoplot +
  scale_y_log10(breaks = c(2, 4, 8, 16, 32)) +
  labs(y = "Time [milliseconds]")
```

## Map functions

* Why `for` loops are good
* Why `map()` functions may be better
* Types of `map()` functions
    * `map()` makes a list
    * `map_lgl()` makes a logical vector
    * `map_int()` makes an integer vector
    * `map_dbl()` makes a double vector
    * `map_chr()` makes a character vector

## Map functions

```{r}
map_dbl(df, mean)
map_dbl(df, median)
map_dbl(df, sd)
```

## Map functions

```{r}
map_dbl(df, mean, na.rm = TRUE)
```

## Map functions

```{r}
df %>%
  map_dbl(mean, na.rm = TRUE)
```
