---
title: "OLS: Interaction terms"
author: |
  | MACS 33001
  | University of Chicago
output: rcfss::cfss_slides
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

library(tidyverse)
library(forcats)
library(broom)
library(modelr)
library(stringr)
library(titanic)
library(rcfss)
library(car)
library(plotly)
library(haven)
library(coefplot)
library(patchwork)

options(digits = 3)
set.seed(1234)
theme_set(theme_minimal(base_size = 22))
```

$$\newcommand{\E}{\mathrm{E}} \newcommand{\Var}{\mathrm{Var}} \newcommand{\Cov}{\mathrm{Cov}} \newcommand{\se}{\text{se}} \newcommand{\Lagr}{\mathcal{L}} \newcommand{\lagr}{\mathcal{l}}$$

## Additive model

$$Y = \beta_0 + \beta_1 X + \beta_2 Z + \epsilon_i$$

## Additive model

```{r sim-linear}
n_sim <- 100

additive <- data_frame(x = runif(n_sim),
                       z = runif(n_sim)) %>%
  mutate(z = ifelse(z > .5, 1, 0),
         y = x + z + rnorm(n_sim))

additive %>%
  add_predictions(lm(y ~ x + z, data = .)) %>%
  ggplot(aes(x, y, color = factor(z))) +
  geom_point() +
  geom_line(aes(y = pred)) +
  theme(legend.position = "none")
```

## Additive model

$$\E[Y] = \beta_0 + \beta_1 X + \beta_2 Z$$

$$\frac{\delta \E[Y]}{\delta X} = \beta_1$$

$$\frac{\delta \E[Y]}{\delta Z} = \beta_2$$

## Multiplicative interaction model

$$Y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ + \epsilon_i$$

* Direct effects
* Constitutive terms
* Interaction term

## Multiplicative interaction model

$$
\begin{align}
\E[Y] & = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 XZ \\
     & = \beta_0 + \beta_2 Z + (\beta_1 + \beta_3 Z) X
\end{align}
$$

$$\frac{\delta \E[Y]}{\delta X} = \beta_1 + \beta_3 Z$$

$$\E[Y] = \beta_0 + \beta_2 Z + \psi_1 X$$

$$
\begin{align}
\E[Y] & = \beta_0 + \beta_1 X + (\beta_2 + \beta_3 X) Z \\
     & = \beta_0 + \beta_1 X + \psi_2 Z
\end{align}
$$

## Multiplicative interaction model

* Conditional impact
* If $Z = 0$, then:
    
    $$
    \begin{align}
    \E[Y] & = \beta_0 + \beta_1 X + \beta_2 (0) + \beta_3 X (0) \\
         & = \beta_0 + \beta_1 X
    \end{align}
    $$
        
* If $X = 0$, then:
    
    $$
    \begin{align}
    \E[Y] & = \beta_0 + \beta_1 (0) + \beta_2 Z + \beta_3 (0) Z \\
         & = \beta_0 + \beta_2 Z
    \end{align}
    $$
* $\psi_1 = \beta_1$ and $\psi_2 = \beta_2$
* $+\beta_3$ and $-\beta_3$
* $\psi_1$ and $\psi_2$

## Conducting inference

* Obtaining estimates of parameters

    $$\hat{\psi}_1 = \hat{\beta}_1 + \hat{\beta}_3 Z$$
    $$\hat{\psi}_2 = \hat{\beta}_2 + \hat{\beta}_3 X$$
    
* Obtaining estimates of standard errors

## Conducting inference

1. $\text{Var}(aX) = a^2 \text{Var}(X)$
1. $\text{Var}(X+Y) = \text{Var}(X) + \text{Var}(Y) + 2 \text{Cov}(X,Y)$
1. $\text{Cov}(X, aY) = a \text{Cov}(X,Y)$

## Conducting inference

$$\widehat{\text{Var}(\hat{\psi}_1}) = \widehat{\text{Var} (\hat{\beta}_1)} +Z^2  \widehat{\text{Var} (\hat{\beta}_3)} + 2 Z \widehat{\text{Cov} (\hat{\beta}_1, \hat{\beta}_3)}$$

$$\widehat{\text{Var}(\hat{\psi}_2}) = \widehat{\text{Var} (\hat{\beta}_2)} + X^2  \widehat{\text{Var} (\hat{\beta}_3)} + 2 X \widehat{\text{Cov} (\hat{\beta}_2, \hat{\beta}_3)}$$

* Depend on $\beta_1$, $\beta_2$, and/or $\beta_3$
* Both also depend on the level/value of the interacted variable

## Two dichtomous covariates

$$Y = \beta_0 + \beta_1 D_1 + \beta_2 D_2 + \beta_3 D_1 D_2 + \epsilon_i$$

$$
\begin{align}
\E[Y | D_1 = 0, D_2 = 0] & = \beta_0 \\
\E[Y | D_1 = 1, D_2 = 0] & = \beta_0 + \beta_1 \\
\E[Y | D_1 = 0, D_2 = 1] & = \beta_0 + \beta_2 \\
\E[Y | D_1 = 1, D_2 = 1] & = \beta_0 + \beta_1 + \beta_2 + \beta_3 \\
\end{align}
$$

## Two dichtomous covariates

```{r sim-two-dich}
two_dich <- data_frame(x = runif(n_sim),
                       z = runif(n_sim)) %>%
  mutate_at(vars(x, z), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 10 + 20 * x - 20 * z + 40 * (x * z) + rnorm(n_sim, 0, 5))
```

```{r sim-two-dich-hist}
ggplot(two_dich, aes(y, color = interaction(x, z))) +
  geom_density() +
  scale_color_discrete(labels = c("D1 = D2 = 0",
                                  "D1 = 1, D2 = 0",
                                  "D1 = 0, D2 = 1",
                                  "D1 = D2 = 1"),
                       guide = guide_legend(nrow = 2)) +
  labs(title = expression(paste("Values of ", Y, " for various combinations of values of ", D[1], ",", D[2])),
       x = "Value of Y",
       color = NULL) +
  theme(legend.position = "bottom")
```

## Two dichtomous covariates

```{r sim-two-dich-box}
ggplot(two_dich, aes(interaction(z, x), y)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(expression("D1 = D2 = 0"),
                                  expression("D1 = 0, D2 = 1"),
                                  expression("D1 = 1, D2 = 0"),
                                  expression("D1 = D2 = 1"))) +
  labs(title = expression(paste("Values of ", Y, " for various combinations of values of ", D[1], ",", D[2])),
       x = NULL,
       y = "Values of Y")
```

## One dichotomous and one continuous covariate

$$Y = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 XD + \epsilon_i$$

$$
\begin{align}
\E[Y | X, D = 0] & = \beta_0 + \beta_1 X \\
\E[Y | X, D = 1] & = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X
\end{align}
$$

----

```{r sim-cont-dich-00}
data_frame(x = runif(n_sim, -2, 2),
           z = runif(n_sim, -2, 2)) %>%
  mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 0 + 20 * x + 0 * z + 0 * (x * z) + rnorm(n_sim, 0, 5)) %>%
  ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
  geom_vline(xintercept = 0, alpha = .3) +
  geom_hline(yintercept = 0, alpha = .3) +
  geom_point() +
  geom_abline(intercept = 0, slope = 20) +
  labs(title = expression({beta[2] == beta[3]} == 0),
       color = "Z",
       shape = "Z") +
  theme(legend.position = "bottom")
```
    
----

```{r sim-cont-dich-01}
data_frame(x = runif(n_sim, -2, 2),
           z = runif(n_sim, -2, 2)) %>%
  mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 0 + 20 * x + 10 * z + 0 * (x * z) + rnorm(n_sim, 0, 5)) %>%
  ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
  geom_vline(xintercept = 0, alpha = .3) +
  geom_hline(yintercept = 0, alpha = .3) +
  geom_point() +
  geom_abline(intercept = 0, slope = 20) +
  geom_abline(intercept = 10, slope = 20, linetype = 2) +
  labs(title = expression(paste({beta[2] != 0}, ",", beta[3] == 0)),
       color = "Z",
       shape = "Z") +
  theme(legend.position = "bottom")
```

----

```{r sim-cont-dich-10}
data_frame(x = runif(n_sim, -2, 2),
           z = runif(n_sim, -2, 2)) %>%
  mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 0 + 20 * x + 0 * z + -40 * (x * z) + rnorm(n_sim, 0, 5)) %>%
  ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
  geom_vline(xintercept = 0, alpha = .3) +
  geom_hline(yintercept = 0, alpha = .3) +
  geom_point() +
  geom_abline(intercept = 0, slope = 20) +
  geom_abline(intercept = 0, slope = -20, linetype = 2) +
  labs(title = expression(paste({beta[2] == 0}, ",", beta[3] != 0)),
       color = "Z",
       shape = "Z") +
  theme(legend.position = "bottom")
```
    
----

```{r sim-cont-dich-11}
data_frame(x = runif(n_sim, -2, 2),
           z = runif(n_sim, -2, 2)) %>%
  mutate_at(vars(z), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 0 + 20 * x + 20 * z + -40 * (x * z) + rnorm(n_sim, 0, 5)) %>%
  ggplot(aes(x, y, color = factor(z), shape = factor(z))) +
  geom_vline(xintercept = 0, alpha = .3) +
  geom_hline(yintercept = 0, alpha = .3) +
  geom_point() +
  geom_abline(intercept = 0, slope = 20) +
  geom_abline(intercept = 20, slope = -20, linetype = 2) +
  labs(title = expression(paste({beta[2] == 0}, ",", beta[3] != 0)),
       color = "Z",
       shape = "Z") +
  theme(legend.position = "bottom")
```

## Two continuous covariates

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2 + \epsilon_i$$

```{r two-cont}
# no interactive effects
data_frame(x = runif(n_sim, 0, 10),
           z = runif(n_sim, 0, 10),
           y = 10 + 10 * x - 10 * z + 0 * x * z) %>%
  plot_ly(x = ~x, y = ~ z, z = ~ y, type = "mesh3d") %>%
  layout(title = "No interactive effects")
```

## Two continuous covariates

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2 + \epsilon_i$$

```{r two-cont-x}
# inteactive effects
data_frame(x = runif(n_sim, 0, 10),
           z = runif(n_sim, 0, 10),
           y = 10 + 10 * x - 10 * z + 10 * x * z) %>%
  plot_ly(x = ~x, y = ~ z, z = ~ y, type = "mesh3d") %>%
  layout(title = "Interactive effects")
```

## Quadratic, cubic, and other polynomial effects

$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon$$

$$\frac{\delta \E[Y]}{\delta X} = \beta_1 + 2 \beta_2 X$$

## Quadratic, cubic, and other polynomial effects

```{r poly-interact, fig.asp = 0.8}
{
  data_frame(x = runif(n_sim, -2, 2)) %>%
    mutate(y = 10 + 10 * x - (50 * x^2) + rnorm(n_sim, 0, 5)) %>%
    add_predictions(lm(y ~ poly(x, 2, raw = TRUE), data = .)) %>%
    ggplot(aes(x, y)) +
    geom_point() +
    geom_line(aes(y = pred)) +
    labs(title = "Quadratic") +
    theme(axis.title = element_blank(),
          axis.text = element_blank())
} +
{
  data_frame(x = runif(n_sim, -2, 2)) %>%
    mutate(y = -500 - 20 * x + 300 * x^2 + rnorm(n_sim, 0, 50)) %>%
    add_predictions(lm(y ~ poly(x, 2, raw = TRUE), data = .)) %>%
    ggplot(aes(x, y)) +
    geom_point() +
    geom_line(aes(y = pred)) +
    labs(title = "Quadratic") +
    theme(axis.title = element_blank(),
          axis.text = element_blank())
} +
{
  data_frame(x = runif(n_sim, -2, 2)) %>%
    mutate(y = 10 + 10 * x - 50 * x^2 + 300 * x^3 + rnorm(n_sim, 0, 300)) %>%
    add_predictions(lm(y ~ poly(x, 3, raw = TRUE), data = .)) %>%
    ggplot(aes(x, y)) +
    geom_point() +
    geom_line(aes(y = pred)) +
    labs(title = "Cubic") +
    theme(axis.title = element_blank(),
          axis.text = element_blank())
}
```

## Higher-order interaction terms

$$
\begin{align}
Y = \beta_0 &+ \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_1 X_2 \\
& + \beta_5 X_1 X_3 + \beta_6 X_2 X_3 + \beta_7 X_1 X_2 X_3 + \epsilon
\end{align}
$$

## Higher-order interaction terms

$$
\begin{align}
Y = \beta_0 &+ \beta_1 X + \beta_2 D_1 + \beta_3 D_2 + \beta_4 X D_1 \\
& + \beta_5 X D_2 + \beta_6 D_1 D_2 + \beta_7 X D_1 D_2 + \epsilon
\end{align}
$$

## Higher-order interaction terms

```{r three-way}
data_frame(x = runif(n_sim, -2, 2),
           d1 = runif(n_sim),
           d2 = runif(n_sim)) %>%
  mutate_at(vars(d1, d2), funs(ifelse(. > .5, 1, 0))) %>%
  mutate(y = 10 + 10 * x + 10 * d1 - 20 * d2 - 20 * x * d1 + 20 * x *d2 - 50 * x * d1 * d2 + rnorm(n_sim, 0, 10)) %>%
  add_predictions(lm(y ~ x * d1 * d2, data = .)) %>%
  ggplot(aes(x, y)) +
  facet_grid(d1 ~ d2, labeller = label_both) +
  geom_point() +
  geom_line(aes(y = pred))
```

## Key rules

* Don't omit the "direct effects"
* Zero should be meaningful
* Rescaling the variables doesn't guarantee statistical significance
* Flexible alternatives
* Interpreting three(+)-way interactions

## Estimating models with multiplicative interactions

````{r nes-data}
# get nes data
nes <- read_dta("../data/nes2008.dta") %>%
  select(obama_therm_post, partyid3, libcon7, libcon7_obama) %>%
  mutate_each(funs(ifelse(is.nan(.), NA, .))) %>%
  rename(ObamaTherm = obama_therm_post,
         RConserv = libcon7,
         ObamaConserv = libcon7_obama) %>%
  mutate(GOP = ifelse(partyid3 == 3, 1, 0)) %>%
  select(-partyid3) %>%
  na.omit
```

* Obama feeling thermometer (`ObamaTherm`)
* `RConserv`
* `ObamaConserv`
* `GOP`

## Obama data

```{r nes-sum}
summary(nes)
```

## Basic linear model

```{r obama-linear}
obama_base <- lm(ObamaTherm ~ RConserv + GOP, data = nes)
tidy(obama_base)
glance(obama_base)
```

## Dichotomous interaction

$$
\begin{align}
\text{Obama} = \beta_0 &+ \beta_1 (\text{RConserv}) \\
& + \beta_2 (\text{GOP})\\
& + \beta_3 (\text{RConserv}) (\text{GOP}) \\
& + \epsilon
\end{align}
$$

## Dichotomous interaction

```{r obama-ideo-x-gop}
obama_ideo_gop <- lm(ObamaTherm ~ RConserv * GOP, data = nes)
tidy(obama_ideo_gop)
glance(obama_ideo_gop)
```

## Dichotomous interaction

* `GOP = 0`

    $$
\begin{align}
E(\text{Obama}) = `r coef(obama_ideo_gop)[["(Intercept)"]]` & `r coef(obama_ideo_gop)[["RConserv"]]` (\text{RConserv})  `r coef(obama_ideo_gop)[["GOP"]]` (0)\\
&  `r coef(obama_ideo_gop)[["RConserv:GOP"]]` (\text{RConserv} \times 0) \\
= `r coef(obama_ideo_gop)[["(Intercept)"]]` & `r coef(obama_ideo_gop)[["RConserv"]]` (\text{RConserv})
\end{align}
    $$

* `GOP = 1`

    $$
\begin{align}
E(\text{Obama}) & = (`r coef(obama_ideo_gop)[["(Intercept)"]]` `r coef(obama_ideo_gop)[["GOP"]]` (1)) + (`r coef(obama_ideo_gop)[["RConserv"]]` `r coef(obama_ideo_gop)[["RConserv:GOP"]]` (\text{RConserv} \times 1)) \\
& = `r coef(obama_ideo_gop)[["(Intercept)"]] + coef(obama_ideo_gop)[["GOP"]]` `r coef(obama_ideo_gop)[["RConserv"]] + coef(obama_ideo_gop)[["RConserv:GOP"]]` (\text{RConserv})
\end{align}
    $$

## Dichotomous interaction

```{r obama-ideo-x-gop-plot}
nes %>%
  add_predictions(obama_ideo_gop) %>%
  ggplot(aes(RConserv, ObamaTherm, color = factor(GOP))) +
  geom_jitter(alpha = .5) +
  geom_line(aes(y = pred)) +
  scale_color_manual(values = c("blue", "red")) +
  labs(x = "Respondent conservatism",
       y = "Obama thermometer score") +
  theme(legend.position = "none")
```

## Separate models

```{r obama-split}
tidy(lm(ObamaTherm ~ RConserv, data = filter(nes, GOP == 0)))
tidy(lm(ObamaTherm ~ RConserv, data = filter(nes, GOP == 1)))
```

## Causal direction

```{r obama-other-direction}
nes %>%
  data_grid(RConserv, GOP) %>%
  add_predictions(obama_ideo_gop) %>%
  spread(GOP, pred) %>%
  mutate(diff = `1` - `0`) %>%
  ggplot(aes(RConserv, diff)) +
  geom_point() +
  labs(title = "Expected Obama thermometer score",
       x = "Respondent conservatism",
       y = "First difference between Republicans\nand non-Republicans")
```

## Calculating standard errors

$$
\begin{align}
\text{Obama} = \beta_0 &+ (\beta_1 + \beta_3 \text{GOP}) (\text{RConserv}) \\
& + \beta_2 (\text{GOP}) + \epsilon \\
 = &\beta_0 + \psi_1 (\text{RConserv}) + \beta_2 (\text{GOP}) + \epsilon
\end{align}
$$

* Point estimate

    ```{r point}
coef(obama_ideo_gop)[["GOP"]] + coef(obama_ideo_gop)[["RConserv:GOP"]]
    ```

* Standard error

    $$\hat{\sigma}_{\hat{\psi}_1} = \sqrt{\widehat{\text{Var}(\hat{\beta}_1)} + (\text{GOP})^2 \widehat{\text{Var}(\hat{\beta_3})} + 2 (\text{GOP}) \widehat{\text{Cov}(\hat{\beta}_1 \hat{\beta}_3)}}$$

    ```{r std-err}
vcov(obama_ideo_gop)

sqrt(vcov(obama_ideo_gop)["RConserv", "RConserv"] +
       (1)^2 * vcov(obama_ideo_gop)["RConserv:GOP", "RConserv:GOP"] +
       2 * 1 * vcov(obama_ideo_gop)["RConserv", "RConserv:GOP"])
    ```

## Conducting inference {.scrollable}

```{r obama-other-dir-std-err}
# function to get point estimates and standard errors
# model - lm object
# mod_var - name of moderating variable in the interaction
instant_effect <- function(model, mod_var){
  # get interaction term name
  int.name <- names(model$coefficients)[[which(str_detect(names(model$coefficients), ":"))]]
  
  marg_var <- str_split(int.name, ":")[[1]][[which(str_split(int.name, ":")[[1]] != mod_var)]]
  
  # store coefficients and covariance matrix
  beta.hat <- coef(model)
  cov <- vcov(model)
  
  # possible set of values for mod_var
  if(class(model)[[1]] == "lm"){
    z <- seq(min(model$model[[mod_var]]), max(model$model[[mod_var]]))
  } else {
    z <- seq(min(model$data[[mod_var]]), max(model$data[[mod_var]]))
  }
  
  # calculate instantaneous effect
  dy.dx <- beta.hat[[marg_var]] + beta.hat[[int.name]] * z
  
  # calculate standard errors for instantaeous effect
  se.dy.dx <- sqrt(cov[marg_var, marg_var] +
                     z^2 * cov[int.name, int.name] +
                     2 * z * cov[marg_var, int.name])
  
  # combine into data frame
  data_frame(z = z,
             dy.dx = dy.dx,
             se = se.dy.dx)
}

# point range plot
instant_effect(obama_ideo_gop, "RConserv") %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of GOP",
       subtitle = "By respondent conservatism",
       x = "Respondent conservatism",
       y = "Estimated marginal effect")

# line plot
instant_effect(obama_ideo_gop, "RConserv") %>%
  ggplot(aes(z, dy.dx)) +
  geom_line() +
  geom_line(aes(y = dy.dx - 1.96 * se), linetype = 2) +
  geom_line(aes(y = dy.dx + 1.96 * se), linetype = 2) +
  geom_hline(yintercept = 0) +
  labs(title = "Marginal effect of GOP",
       subtitle = "By respondent conservatism",
       x = "Respondent conservatism",
       y = "Estimated marginal effect")
```

## Hypothesis testing {.scrollable}

```{r hypo-test, echo = TRUE}
linearHypothesis(obama_ideo_gop, "RConserv + RConserv:GOP")
```

```{r hypo-test-rconserv7, echo = TRUE}
linearHypothesis(obama_ideo_gop, "GOP + 7 * RConserv:GOP")
```

## Continuous interaction

$$
\begin{align}
\text{Obama} = \beta_0 &+ \beta_1 (\text{RConserv}) \\
& + \beta_2 (\text{ObamaConserv})\\
& + \beta_3 (\text{RConserv}) (\text{ObamaConserv}) \\
& + \epsilon
\end{align}
$$

## Continuous interaction

```{r obama-two-cont}
obama_ideo2 <- lm(ObamaTherm ~ RConserv * ObamaConserv, data = nes)
tidy(obama_ideo2)
glance(obama_ideo2)
```

## Continuous interaction {.scrollable}

```{r obama-two-marg-effect}
instant_effect(obama_ideo2, "RConserv") %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of Obama's conservatism",
       subtitle = "By respondent conservatism",
       x = "Respondent conservatism",
       y = "Estimated marginal effect")

instant_effect(obama_ideo2, "ObamaConserv") %>%
  ggplot(aes(z, dy.dx,
             ymin = dy.dx - 1.96 * se,
             ymax = dy.dx + 1.96 * se)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Marginal effect of respondent's conservatism",
       subtitle = "By Obama conservatism",
       x = "Obama conservatism",
       y = "Estimated marginal effect")
```

## Predicted values plots

```{r obama-heatmap}
nes %>%
  data_grid(RConserv, ObamaConserv) %>%
  add_predictions(obama_ideo2) %>%
  ggplot(aes(RConserv, ObamaConserv, z = pred, fill = pred)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(midpoint = 50) +
  geom_contour() +
  labs(title = "Expected Obama thermometer score",
       x = "Respondent conservatism",
       y = "Obama conservatism",
       fill = "Prediction")
```

## Predicted values plots

```{r obama-3d}
nes %>%
  data_grid(RConserv, ObamaConserv) %>%
  add_predictions(obama_ideo2)  %>%
  spread(ObamaConserv, pred) %>%
  select(-RConserv) %>%
  as.matrix %>%
  plot_ly(x = ~ 1:7, y = ~ 1:7, z = .) %>%
  add_surface() %>%
  layout(title = "Predicted values for Obama feeling thermometer",
         scene = list(
           xaxis = list(
             title = "Respondent conservatism"
           ),
           yaxis = list(
             title = "Obama conservatism"
           ),
           zaxis = list(
             title = "Predictions"
           )
         )
  )
```
